Rclean and Automatng Code Cleaning in R

### 2020-03-12 ropensci/rclean_technote.Rmd
Edits technote and adds references

Rclean: Automated Tools for Cleaning R Code
===========================================

> Leave the code cleaner than you found it. 
-- R.C. Martin in *Clean Code*

@@

As you can see, the provenance provides a network representation of
relationships among functions and objects. This is very powerful
because we can now apply algorithms analyze the R script with respect
to our results. The main function of `Rclean`, called `clean`, takes
the provenance and applies a depth first network search algorithm to
determine the pathways leading from inputs to outputs, excluding any
objects or lines of code that do not fall along that pathway and are,
therefore, not necessary to produce the desired set of results. As
demonstrated in the next section, through this process, we can produce
a new script with the minimal code necessary to generate the output we
want.

@@

The following is an example of a script that has some "complications."
As you can see, although the script is not extremely long or
complicated, it's difficult enough to make it frustrating to visualize
it in its entirety and pick through it.

```{R long-setup, echo = FALSE, eval = TRUE, results = "hide"}

@@

```{R long-script, echo = FALSE, eval = FALSE}

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit_sqrt_A`, which is an analysis that is
relevant to some product. Not only do we want to double check the
results, we also want to use the code again for another purpose, such
as creating a plot of the patterns supported by the test. Manually
tracing through our code for all the variables used in the test and
finding all of the lines that were used to prepare them for the
analysis would be annoying and difficult, especially given the fact
that we have used "x" as a prefix for multiple unrelated objects in
the script. Instead, we can easily do this automatically with
`Rclean`.

@@

Currently, because of the data provenance that is used, libraries can
not be isolated directly during the cleaning process. So, the
`get_libs` function provides a way to detect the libraries for a given
script. Just supply a file path and `get_libs` will return the
libraries that are called by that script.

@@

## Software Availability

The software is currently hosted on Github, and we recommend using the
`devtools` library [@R-devtools] to install directly from the
repository (https://github.com/ROpenSci/Rclean). The package is
open-source and welcomes contributions. Please visit the repository
page to report issues, request features or provide other feedback.

@@

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as processing information from comments, parsing
control statements and replicating random processes. However, using
retrospective provenance comes at a cost. In order to gather it, the
script needs to be executed. When scripts are computationally
intensive or contain bugs that stop execution, then retrospective
provenance can not be obtained for part or all of the code. Some work
has already been done in the direction of implementing retrospective
provenance for code cleaning in R (see
http://end-to-end-provenance.github.io).

>>>>

---
title: Rclean and Automatng Code Cleaning in R
author: M.K. Lau
bibliography: refs.bib
---

> Leave the code cleaner than you found it.

-- R.C. Martin in *Clean Code*

@@

The provenance provides a network representation of relationships
among functions and objects. This is very powerful because we can now
apply algorithms analyze the R script with respect to our results. The
main function of `Rclean`, called `clean`, takes the provenance and
applies a depth first network search algorithm to determine the
pathways leading from inputs to outputs, excluding any objects or
lines of code that do not fall along that pathway and are, therefore,
not necessary to produce the desired set of results. As demonstrated
in the next section, through this process, we can produce a new script
with the minimal code necessary to generate the output we want.

@@

Analytical scripts that have not been refactored are often both long
and complicated. However, a script doesn't need to be long to be
complicated. The following example script presents some challenges
such that even though it's not a long script, picking through it to
get a result would likely prove to be frustrating.

```{R long-setup, echo = FALSE, eval = TRUE}

@@

```{R long-script, echo = TRUE, eval = FALSE}

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit_sqrt_A`, which is an analysis that is
the fitted model object for an analysis. We might want to double check
the results, and we also might need to use the code again for another
purpose, such as creating a plot of the patterns supported by the
test. Manually tracing through our code for all the variables used in
the test and finding all of the lines that were used to prepare them
for the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. However, `Rclean` can do this easily via the `clean`.

@@

Currently, libraries can not be isolated directly during the cleaning
process. So, the `get_libs` function provides a way to detect the
libraries for a given script. We just need to supply a file path and
`get_libs` will return the libraries that are called by that script.

@@

## Software Availability and Installation

Thanks to the efforts and helpful feedback from the ROpenSci
Community, the package has recently passed software review and we have
also recently published an article in the Journal of Statistical
Software. The package is hosted through the ROpenSci organization on
Github, and we recommend using the `devtools` library [@R-devtools] to
install directly from the repository (https://github.com/ROpenSci/Rclean).

```{R install}
library(devtools)
install_github("ROpenSci/Rclean")

@@

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as libraries that are actually used by the code,
processing comments, parsing control statements and replicating random
processes. Using retrospective provenance comes at a cost. In order to
gather it, the script needs to be executed. When scripts are
computationally intensive or contain bugs that stop execution, then
retrospective provenance can not be obtained for part or all of the
code. Some work has already been done in the direction of implementing
retrospective provenance for code cleaning in R (see
http://end-to-end-provenance.github.io). Although such costs may
present challenges, combining prospective and retrospective provenance
methods could provide a powerful and flexible solution.
